Handling Recursion in Generic Programming Using Closed Type Families

### 2018-04-26 lncs-paper.tex
Remove appendices

\begin{document}
%
\title{Handling Recursion in Generic Programming Using Closed Type Families}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Anonymous%First Author\inst{1} \and
%Second Author\inst{2}
}
%
%\authorrunning{F. Author and S. Author}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%~ \institute{Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany\\
%~ \email{lncs@springer.com} \and
%~ ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%~ \email{abc@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express a recursive structure. Some approaches based on sums of products, however, do not use a fixed point. Those views therefore allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues when it needs to deal with recursion. A known and unwelcome solution is the use of overlapping instances. We present an approach that uses closed type families to eliminate the need of overlap for handling recursion. Moreover, we show that our idiom allows for families of mutually recursive datatypes.

@@

A classical way to generically describe a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}. De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types. They call their view SOP which stands for a ``sum of products''~--- this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
%, and \emph{GADTs}~\cite{Schrijvers2009}. -- very old extension, 2003 approx.
Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.

@@

One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed point approach. This may complicate the whole framework significantly. Besides, such a decision may lead to extra conversions between the generic views.

Another known solution uses overlapping instances. This usually unwelcome Haskell extension complicates reasoning about the semantics of code. In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances. Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
The overlap problem also strikes in the security setting when a code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way. Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
% -- this should go to Realted Work (if stays in the paper at all)

@@

The function \texttt{gshow} involves meta-information provided by \textsf{generics-sop}'s class \texttt{HasDatatypeInfo} to show a datatype constructor's and its record fields' names. The \textsf{generics-sop} library has features for deriving this meta-information automatically.
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String
\vs
}
We define \texttt{gshow} mutually recursive with \texttt{caseShow}. The function is constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types. The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in Appendix~1.

@@

The definition of \texttt{toFirstCtx} can now be given using the defined functions \texttt{appInjCtx} and \texttt{fromFstRec}. The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurence in the given datatype's product list with type \texttt{Hole}. Its definition is straightforward, and is omitted here.
\texttt{
\vs
\indent toFirstCtx \ann \fa fam c a. (Generic a, ToFirstCtx fam a)\\
\indent\ind\ind\s\s\s \Ra a \ra Context fam a\\
\indent toFirstCtx t = toFirstCtxNS @'F \$ unSOP \$ from t
\vs
\indent toFirstCtxNS \ann \fa n fam a xss. ToFirstCtx' fam a n xss\\
\indent\ind\ind\ind\s \Ra NS (NP I) xss \ra Context fam a\\
\indent toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns\\
\indent toFirstCtxNS (Z (np \ann NP I xs))\\
\indent\s\s = appInjCtx @(CtxConsN (CtxCode fam a) n) \$\\
\indent\ind\ind I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
\vs
\indent type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
\vs
}
The \texttt{ToFirstCtx'} constraint here is a complex constraint for use of \texttt{appInjCtx} and \texttt{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints. Its definition is shown in Appendix~2. We omit this here for reasons of space.

@@

Over the years, there has been a lot of works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated. Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approachs and provide their detailed comparison in different aspects. There are several generic views that use forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}. And there are a number of approachs that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation. The SOP view~\cite{VriLoeh2014} which we use to demonstrate our technique is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype. This approach uses heterogeneous lists of types to encode sums and products in the generic representation.

@@

It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approachs to generic programming was considered as unsuitable for this task. There are some approachs that address this problem by means of local overlaps, where it needs to access the points of recursion. We have developed the technique that allows us to define generic functions that treat recursion without its explicit encoding and overlap. This makes use of closed type families that are the recent Haskell extension.

We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface, and supports families of mutually recursive datatypes. Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approachs is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.

\section*{Appendix 1}

In this appendix, we show the implementation of the generic function \texttt{gshow}, which uses \texttt{caseShow}, described in Section~\ref{sec:gshow}. The \texttt{gshow} function is originally defined by De Vries and L\"{o}h in the \textsf{basic-sop} package~\cite{basic-sop}. Their implementation has the issue that for recursive datatypes whose recursive nodes should be displayed within parentheses, the function acts incorrectly due to the SOP view's limitations on expressing recursion. We improve the implementation from \textsf{basic-sop} with our approach to handling recursion.

The \texttt{gshow} function makes use of metadata provided by the \textsf{generics-sop} library through the class \texttt{HasDatatypeInfo}. That class delivers generic functions for handling meta-information which \textsf{generics-sop} defines separate from the main generic representation of a datatype. The definition of \texttt{gshow} using those functions is as follows:
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String\\
\indent gshow t = gshow' @a (constructorInfo \$ datatypeInfo \$ Proxy @a)\\
\indent\ind\ind\ind\ind\ind \$ from t
\vs
}
The functions from \textsf{generics-sop} use a proxy to fix a type value, where we use \textsf{TypeApplications}---a later language extension than those the library relies on. The auxiliary function \texttt{gshow'}, which works on the metadata encoding and generic representation of a datatype, uses \textsf{generics-sop}'s combinators for collapsing and mapping:
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\bf Processing constructors
\vs
\tt goConstructor \ann\hs \fa a xs. All (CaseRecShow a) xs\\
\tt\ind\ind\ind\s\s \Ra ConstructorInfo xs \ra NP I xs \ra K String xs\\
\tt goConstructor (Constructor n) args = K \$ unwords (n : args')\\
\tt\s\s \textbf{where}\\
\tt\ind args' \ann [String]\\
\tt\ind args' = hcollapse \$ hcmap (p @a) (K . caseShow @a . unI) args\\
\tt goConstructor (Record n ns) args\\
\tt\ind = K \$ n ++ " \{" ++ intercalate ", " args' ++ "\}"\\
\tt\s\s \textbf{where}\\
\tt\ind args' \ann [String]\\
\tt\ind args' = hcollapse \$ hczipWith (p @a) (goField @a) ns args\\
\tt goConstructor (Infix n \_ \_) (I arg1 :* I arg2 :* Nil)\\
\tt\ind = K \$ caseShow @a arg1 ++ " " ++ n ++ " " ++ caseShow @a arg2
\vs
\tt p \ann Proxy (CaseRecShow a)\\
\tt p = Proxy
\vspace{0.3cm}\\
\bf Processing record fields
\vs
\tt goField \ann\hs \fa a x. (CaseRecShow a) x\\
\tt\ind\ind \Ra FieldInfo x \ra I x \ra K String x\\
\tt goField (FieldInfo n) (I y)\\
\tt\ind = K \$ n ++ " = " ++ caseShow @a y
\end{tabular}
\caption{Auxiliary functions for implementing \texttt{gshow}.}
\label{fig:cons-fields}
\end{figure}
\texttt{
\vs
\indent gshow' \ann\hs \fa a xss. All2 (CaseRecShow a) xss\\
\indent\ind\s\s\s \Ra NP ConstructorInfo xss \ra SOP I xss \ra String\\
\indent gshow' cs (SOP sop)\\
\indent\ind = hcollapse \$ hczipWith allp (goConstructor @a) cs sop\\
\indent\s\s \textbf{where}\\
\indent\ind allp = Proxy @(All (CaseRecShow a))
\vs
}
When a sum structure is actually homogeneous (i.~e., has type \texttt{NS~(K~a)~xs}), it can be collapsed to a single component of type \texttt{a}. The respective instantiation of the function \texttt{hcollapse} that generalizes collapsing homogeneous structures is
\texttt{
\vs
\indent hcollapse  \ann NS  (K a) xs  \ra  a
\vs
}
The \texttt{hczipWith} function is generalized \texttt{zipWith} that operates with a constrained function on heterogeneous structures, where the proxy fixes the constraint:
\texttt{
\vs
\indent hczipWith \ann\hs All c xs\\
\indent\ind\ind\s\s \Ra proxy c \ra (\fa a. c a \Ra f a \ra g a \ra h a)\\
\indent\ind\ind\s\s \ra NP f xs \ra NS g xs \ra NS h xs
\vs
}
In Figure~\ref{fig:cons-fields}, we show functions that process constructors and record fields obtaining their names from proper constructors of metadata and make use of \texttt{caseShow}. The function \texttt{goConstructor} repeats in general the shape of \texttt{gshow'}, where the \texttt{hcmap} function generalizes \texttt{map} for \texttt{NP}, the return type of \texttt{hczipWith} on two products becomes \texttt{NP}, and on two cases, \texttt{hcollapse} collapses the result \texttt{NP} to a list of strings.

\section*{Appendix 2}

Here, we give the full definition of the \texttt{ToFirstCxt'} constraint using definitions from Section~\ref{sec:generic-zipper}.
\texttt{
\vs
\indent type FstCtx n fam xs = Proxy n ': FstRecToHole fam xs
\vs
\indent class ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
\indent \s\s\Ra CtxConsNInj fam a xs n\\
\indent instance ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
\indent \s\s\Ra CtxConsNInj fam a xs n
\vs
\indent class All (ProofF f (FstCtx n fam xs)) yss\\
\indent \s\s\Ra CtxAllProofF f fam yss xs n\\
\indent instance All (ProofF f (FstCtx n fam xs)) yss\\
\indent \s\s\Ra CtxAllProofF f fam yss xs n
\vs
\indent class FromFstRec (FstRecToHole fam xs) xs\\
\indent \s\s\Ra FromFstRec' fam xs\\
\indent instance FromFstRec (FstRecToHole fam xs) xs\\
\indent \s\s\Ra FromFstRec' fam xs
\vs
\indent type family AllNum (c \ann k \ra ConsN \ra Constraint)\\
\indent \ind\ind\ind\ind\s\s\s (xs \ann [k]) (n \ann ConsN) \ann Constraint where\\
\indent \s\s AllNum c '[]\ind\s\s\s n = ()\\
\indent \s\s AllNum c (x ': xs) n = (c x n, AllNum c xs ('N n))
\vs
\indent type ToFirstCtx' fam a n xss\\
\indent \ind = (SListI (CtxCode fam a),\\
\indent \ind\s\s\s All (FromFstRec' fam) xss,\\
\indent \ind\s\s\s AllNum (CtxConsNInj fam a) xss n,\\
\indent \ind\s\s\s AllNum (CtxAllProofF\\
\indent \ind\ind\ind\ind\s (Inj (NP I) (CtxCode fam a))\\
\indent \ind\ind\ind\ind\s fam (CtxCode fam a)) xss n)
}

>>>>

\begin{document}
%
\title{Handling Recursion in Generic Programming Using Closed Type Families}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Anna Bolotina\inst{1} \and
Artem Pelenitsyn\inst{2}
}
%
%\authorrunning{F. Author and S. Author}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Southern Federal University, Rostov-on-Don, Russia\\
\email{bolotina@sfedu.ru} \and
Czech Technical University in Prague, Prague, Czech Republic\\
\email{pelenart@fit.cvut.cz}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express a recursive structure. Some approaches based on sums of products, however, do not use a fixed point. Those views therefore allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues when it needs to deal with recursion. A known and unwelcome solution is the use of overlapping instances. We present an approach that uses closed type families to eliminate the need of overlap for handling recursion. Moreover, we show that our idiom allows for families of mutually recursive datatypes.

@@

A classical way to generically express a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}. De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types. They call their view SOP which stands for a ``sum of products''---this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
%, and \emph{GADTs}~\cite{Schrijvers2009}. -- very old extension, 2003 approx.
Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.

@@

One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed-point approach. This may complicate the whole framework significantly. Besides, such a decision may lead to extra conversions between the generic views.

Another known solution uses overlapping instances. This usually unwelcome Haskell extension complicates reasoning about the semantics of code. In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances. Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way. Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
% -- this should go to Realted Work (if stays in the paper at all)

@@

The function \texttt{gshow} involves meta-information provided by \textsf{generics-sop}'s class \texttt{HasDatatypeInfo} to show a datatype constructor's and its record fields' names. The \textsf{generics-sop} library has features for deriving this meta-information automatically.
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String
\vs
}
We define \texttt{gshow} mutually recursive with \texttt{caseShow}. The function is constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types. The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.

@@

The definition of \texttt{toFirstCtx} can now be given using the defined functions \texttt{appInjCtx} and \texttt{fromFstRec}. The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurrence in the given datatype's product list with type \texttt{Hole}. Its definition is straightforward, and is omitted here.
\texttt{
\vs
\indent toFirstCtx \ann \fa fam c a. (Generic a, ToFirstCtx fam a)\\
\indent\ind\ind\s\s\s \Ra a \ra Context fam a\\
\indent toFirstCtx t = toFirstCtxNS @'F \$ unSOP \$ from t
\vs
\indent toFirstCtxNS \ann \fa n fam a xss. ToFirstCtx' fam a n xss\\
\indent\ind\ind\ind\s \Ra NS (NP I) xss \ra Context fam a\\
\indent toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns\\
\indent toFirstCtxNS (Z (np \ann NP I xs))\\
\indent\s\s = appInjCtx @(CtxConsN (CtxCode fam a) n) \$\\
\indent\ind\ind I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
\vs
\indent type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
\vs
}
The \texttt{ToFirstCtx'} constraint here is a complex constraint for use of \texttt{appInjCtx} and \texttt{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints. Its definition is given in the extended version of the paper in Technical Report. We omit this here for reasons of space.

@@

Over the years, there has been a lot of works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated. Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in different aspects. There are several generic views that use forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}. And there are a number of approaches that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation. The SOP view~\cite{VriLoeh2014} which we use to demonstrate our technique is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype. This approach uses heterogeneous lists of types to encode sums and products in the generic representation.

@@

It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approaches to generic programming was considered as unsuitable for this task. There are some approaches that address this problem by means of local overlaps, where it needs to access the points of recursion. We have developed the technique that allows us to define generic functions that treat recursion without its explicit encoding and overlap. This makes use of closed type families that are the recent Haskell extension.

We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface, and supports families of mutually recursive datatypes. Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approaches is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.
